A Hitchhiker's Guide to Reproducible Research in R

### 2021-07-14 manuscript.Rmd
final comments

---
title: A Hitchhiker's Guide to Reproducible Research in R
author:
  - name: Aaron Peikert
    affil: 1, *
    orcid: 0000-0001-7813-818X
  - name: Caspar J. Van Lissa
    affil: 2, 3
    orcid: 0000-0002-0808-5024
  - name: Andreas M. Brandmaier
    affil: 1, 4
    orcid: 0000-0001-8765-6982
affiliation:
  - num: 1
    address: |
      Center for Lifespan Psychology---Max Planck Institute for Human Development
      Lentzeallee 94, 14195 Berlin, Germany
  - num: 2
    address: |
      Department of Methodology & Statistics---Utrecht University faculty of Social and Behavioral Sciences, Utrecht, Netherlands
  - num: 3
    address: |
      Open Science Community Utrecht, Utrecht, Netherlands
  - num: 4
    address: |
      Max Planck UCL Centre for Computational Psychiatry and Ageing Research
      Berlin, Germany and London, UK
# firstnote to eighthnote
correspondence: |
  peikert@mpib-berlin.mpg.de
journal: psych
type: tutorial
status: submit
bibliography: temp.bib
simplesummary: |
  A Simple summary goes here.
abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.Rmd'))))`"
keywords: |
  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
  to the article, yet reasonably common within the subject discipline.).
acknowledgement: |
  The authors received no financial support for the research, authorship, and/or publication of this article.
authorcontributions: |
  Aaron Peikert took the lead in writing and provided the initial draft of the manuscript.
  Andreas Brandmaier and Caspar J. Van Lissa contributed further ideas, critical feedback, and revisions of the original manuscript.
  Furthermore, we would like to thank Maximilian Stefan Ernst (not an author) for his contributions to the code for the simulation study.
conflictsofinterest: |
  The authors declare no conflict of interest.
abbreviations:
  - short: PAC
    long: Preregistration as 
  - short: Gb
    long: Gigabyte
  - short: Kb
    long: Kilobyte
  - short: CRAN
    long: Comprehensive R Archive Network
repro:
  packages:
    - tidyverse
    - usethis
    - gert
    - aaronpeikert/repro@7bfaf98
    - here
    - rstudio/webshot2@f62e743
    - targets
    - renv
    - slider
    - patchwork
    - knitr
    - pander
    - lavaan
    - furrr
    - future.batchtools
    - rticles
    - moments
    - report
  scripts:
    - R/simulation.R
    - R/simulation_funs.R
    - R/link.R
  data:
    - data/simulation_results.csv
output:
  bookdown::markdown_document2:
    base_format: rticles::mdpi_article
header-includes:
   - \usepackage[labelformat=empty]{caption}
   - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, numbers=left}
---

@@

The above mentioned tools require much expertise that is beyond the traditional curriculum of researchers.
To aid researcher adopting these tools we rely on or have developed helpers that wrap more complicated (and powerful) tools.
These wrappers use standard tools but provide feedback for the user about what the computer does and what the user should do in layman's terms and are accessible from within R.
We hope this makes reproducibility tools more accessible by enabling beginner-level users to detect their systems state accurately and act correspondingly [@parasuramanAutomationHumanPerformance2018, Chapter 8: "Automation and Situation Awareness"].
These alternative tools are merely an assistance system, and as users get more and more comfortable, they can use the underlying tools directly to solve more complex problems.

@@

Some projects, however, are not as easy to make reproducible.
As explained above, `repro` is merely a simplified interface that guarantees a high level of reproducibility for simple projects.
These simplifications are possible because of two restrictions.
Users who ask themselves either "How can I install software that is not an R package in the Docker image?" or "How can I express complex dependencies between files, e.g. raw data turned into preprocessed data required by analysis?" need to be aware of these restriction and require a deeper understanding of the inner workings of `repro`.
All other users may safely skip this section and may return to it if they become limited by restriction.

The metadata the `repro::automate()` function relies on can only express dependencies R packages for the Dockerfile and only trivial dependencies (in the form of "file must exist") for the Makefile. 
The first limitation means that users must rely on software that is either already provided by the base Dockerimage "rocker/verse" or the R packages they list in the metadata.
Other software that users might need like other programming languages, not already installed latex packages etc. have to be added manually.
We plan to add support for commonly used ways to install software beyond R packages via the metadata and `repro::automate()` e.g. for system libraries (via `apt` the Ubuntu package manager), Latex packages (via `tlmgr` the Tex Live package manager), python packages (via `pip` the python package manager).
The second limitation is about dependencies.
Make can represent complex dependencies like in this example: A depends on B, which in turn depends on C and D.
If B is missing in this example, Make would know how to recreate it from C and D.
These dependencies and especially how these dependencies should be resolved are difficult to represent in the metadata.
Users therefore have to either "flatten" the dependency structure, by simply stating that A depends on B, C and D therefore leaving out important information or express the dependencies directly within the Makefile.

The following section explains how to overcome these limitation while still relying on the automatons afforded by `repro`.
Lifting these restrictions requires the user to interact more directly with Make or Docker.
To satisfy more complicated requirements users need to understand how repro utilizes Make and Docker internally.

@@

With this approach, users of `repro` can not only build complex software environments but they can also implement complex file dependencies.
The standard `repro` metadata only makes sure that all dependencies are available but does not allow you to specify custom recipes for them in the metadata.
If you can formulate the creation of dependencies in terms of computational steps, e.g. the file `data/clean.csv` is created from `data/raw.csv` by script `R/preprocess.R`, you should include these in the Makefile.
The Makefile that `repro` creates is only an template and you are free to change it.
However, make sure to never remove the following two lines:

@@

The file `.repro/Makefile_Docker` does again contain a rather complicated template which you could but usually should not modify.
This Makefile coordinates the interplay between Make and Docker and contains targets for building (with `make docker`) and saving (with `make save-docker`) the Docker image.
Additionally, it provides facilities to execute commands within the container.
If you write computational recipe for a target it will be by default evaluated using the locally installed software.
To instead evaluate commands inside the Docker image, you should wrap them in `$(RUN1) command $(RUN2)`, like in this example, which is identical to the first Make example we gave above except for the addition of `$(RUN1)` and `$(RUN2)` in l. 2:
```bash
simulated_data.csv: R/simulate.R
  $(RUN1) Rscript -e 'source("R/simulate.R")' $(RUN1)
```

If users execute this in the terminal:

```bash
make simulated_data.csv
```
It behaves exactly as the in the first Make example, the script `R/simulate.R` is run using the locally installed R.
Because this translates simply to:

@@

it is evaluated within the Docker container using the software within it and not the locally installed R version:

@@

At the heart of a PAC lies the planned analysis.
We suggest writing a function that captures the full process of testing a research question that receives a dataset as input.
For the hypothetical research question, we could for example plan to conduct a simple Student's t-test [@studentProbableErrorMean1908] with Welch's correction [@welchGeneralizationStudentProblem1947] an the average of machiavillism items of the Short Dard Triad [SD3, @Jones2013] between the gender groups.
As an example of a data dependent decisions we included a rank transformation if the skewness greater than 1, converting the t-test to a supposedly more robust Mann–Whitney–Wilcoxon test.[^caution-mww]

[^caution-mww]: The Mann–Whitney–Wilcoxon is **not** more robust than the Welch corrected t-test for reasonably large sample sizes. The interested reader can use the provided code for the simulation to verify that the t-test provides unbiased effect sizes but the Mann–Whitney–Wilcoxon overestimates effect size with increasing sample size and skewness.

This function is one possible translation of the verbal descriptions:

@@

One point of note about the analytical strategy and the function implementing it is that there are several translations into code that would fit the descriptions but would produce different results.
One example would be using the function `wilcox.test` instead of the combinations of `rank` and `t.test`.
Both are valid implementation of Mann–Whitney–Wilcoxon test but the first assumes equal variance while the second applies Welch's correction and hence is robust even with unequal variances [@zimmermanRankTransformationsPower1993].
Mentioning every such minute implementation detail is would result in overly verbose preregistrations but does change the outcome.

We also want to caution researchers to not use the here shown procedure.
We merely choose it for didactic purposes because it may be familiar to many researchers because the t-test and Mann–Whitney–Wilcoxon test are arguably the most often used hypothesis test [@fagerlandTtestsNonparametricTests2012; @hortonStatisticalMethodsJournal2005 reports that 26% of all studies employ a t-test and 27% employ a rank based alternative in the New England Journal of Medicine in 2005].
This analytical strategie is suboptimal with regard to several criteria (untested assumption of measurement invariance [@putnickMeasurementInvarianceConventions2016], underestimation of effect size in presence of measurement error [@frostCorrectingRegressionDilution2000], overestimation of effect size for highly skewed distributions [@stonehouseRobustnessTestsCombined1998]);

To use this function one needs a compatible dataset.
At the time of preregistration there is of course no actual data available.
We therefore simulate a dataset.

```{r simulate_data}
```

We decided to draw non-normal data to test how well our procedure performed under non-optimal situations.
Specifically, we draw from a $\chi^2$-distribution, where in one group we add difference that conforms to a standardized effect size.
This simulated dataset fits perfectly into our planned analysis:

@@

These results are based on simulated data:

```{r, results='asis'}
report::report(results$test)
```

@@

@@

>>>>

---
title: A Hitchhiker's Guide to Reproducible Research in R
author:
  - name: Aaron Peikert
    affil: 1, *
    orcid: 0000-0001-7813-818X
  - name: Caspar J. Van Lissa
    affil: 2, 3
    orcid: 0000-0002-0808-5024
  - name: Andreas M. Brandmaier
    affil: 1, 4
    orcid: 0000-0001-8765-6982
affiliation:
  - num: 1
    address: |
      Center for Lifespan Psychology---Max Planck Institute for Human Development
      Lentzeallee 94, 14195 Berlin, Germany
  - num: 2
    address: |
      Department of Methodology & Statistics---Utrecht University faculty of Social and Behavioral Sciences, Utrecht, Netherlands
  - num: 3
    address: |
      Open Science Community Utrecht, Utrecht, Netherlands
  - num: 4
    address: |
      Max Planck UCL Centre for Computational Psychiatry and Ageing Research
      Berlin, Germany and London, UK
# firstnote to eighthnote
correspondence: |
  peikert@mpib-berlin.mpg.de
journal: psych
type: tutorial
status: submit
bibliography: temp.bib
simplesummary: |
  A Simple summary goes here.
abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.Rmd'))))`"
keywords: |
  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
  to the article, yet reasonably common within the subject discipline.).
acknowledgement: |
  The authors received no financial support for the research, authorship, and/or publication of this article.
authorcontributions: |
  Aaron Peikert took the lead in writing and provided the initial draft of the manuscript.
  Andreas Brandmaier and Caspar J. Van Lissa contributed further ideas, critical feedback, and revisions of the original manuscript.
  Furthermore, we would like to thank Maximilian Stefan Ernst (not an author) for his contributions to the code for the simulation study.
conflictsofinterest: |
  The authors declare no conflict of interest.
abbreviations:
  - short: PAC
    long: Preregistration as 
  - short: Gb
    long: Gigabyte
  - short: Kb
    long: Kilobyte
  - short: GUI
    long: Graphical User Interface
  - short: CLI
    long: Comand Line Interface
  - short: CRAN
    long: Comprehensive R Archive Network
repro:
  packages:
    - tidyverse
    - usethis
    - gert
    - aaronpeikert/repro@7bfaf98
    - here
    - rstudio/webshot2@f62e743
    - targets
    - renv
    - slider
    - patchwork
    - knitr
    - pander
    - lavaan
    - furrr
    - future.batchtools
    - rticles
    - moments
    - report
  scripts:
    - R/simulation.R
    - R/simulation_funs.R
    - R/link.R
  data:
    - data/simulation_results.csv
output:
  bookdown::markdown_document2:
    base_format: rticles::mdpi_article
header-includes:
   - \usepackage[labelformat=empty]{caption}
   - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}, numbers=left}
---

@@

A major barrier to adoption of the above tools for researchers is that they are typically used via a command line interface (CLI).
The CLI predates and coexists with graphical user interfaces (GUIs), which many users (including the authors) prefer over the text-based interactions of a CLI for most tasks.
A GUI has the apparent disadvantage for reproducibility that it requires a user who clicks on the correct elements (such as buttons or menu items) in the correct order.<!--AB not sure about the entire paragraph about CLI vs GUI here -  can we shorten this entire paragraph?-->
Depending on the number of steps taken, this is not only much effort but also error-prone.
This tutorial seeks a compromise.
We use simple GUI actions to create a reproducible workflow template, but the reproduction of such workflow does not rely on manual steps. <!--AB I modified this sentence - but I am not sure what we are saying here with the "manual steps"-->
At several points, we rely on or have developed a CLI that wraps a more complicated (and powerful) CLI.
These wrappers use standard CLI tools but provide feedback for the user about what the computer does and what the user should do in layman's terms and are accessible from within R.
We hope this makes reproducibility tools more accessible by enabling beginner-level users to detect their systems state accurately and act correspondingly [@parasuramanAutomationHumanPerformance2018, Chapter 8: "Automation and Situation Awareness"].
These alternative tools are merely an assistance system, and as users get more and more comfortable, they can use the underlying tools directly to solve more complex problems.

@@

Some projects, however, are not as easy to make reproducible.
As explained above, repro is merely a simplified interface that guarantuees a high level of reproducibility for simple projects.
These simplifications are possible because of two restrictions.
First, the project can only rely on R packages for the Dockerfile and second, it can have no multilayered preprocessing for the Makefile.<!--AB: please explain both points in more detail-->
Lifting these restrictions requires the user to interact more directly with Make or Docker.
To satisfy more complicated requirements users need to understand how repro utilizes Make and Docker internally.

@@

With this approach, users of `repro` can not only build complex software environments but they can also implement complex file dependencies.
The standard `repro` metadata only makes sure that all dependencies are available but does not allow you to specify custom recipes for them in the metadata.
If you can formulate the creation of dependencies in terms of computational steps you should include these in the Makefile.<!--AB I do not understand this sentence-->
The Makefile that `repro` creates is only an template and you are free to change it.
However, make sure to never remove the following two lines:

@@

The file `.repro/Makefile_Docker` does again contain a rather complicated template which you could but usually should not modify.
This Makefile coordinates the interplay between Make and Docker and contains targets for building (with `make docker`) and saving (with `make save-docker`) the Docker image.
Additionally, it provides facilities to execute commands within the container.
If you write computational recipe for a target it will be by default evaluated using the locally installed software.
To instead evaluate commands inside the Docker image, you should wrap them in `$(RUN1) command $(RUN2)`, like in this example:<!--AB please explain this hypothetical example in more detail, so that we can follow along-->

```bash
simulated_data.csv: R/simulate.R
  $(RUN1) Rscript -e 'source("R/simulate.R")' $(RUN1)
```

If users execute this in the terminal

```bash
make simulated_data.csv
```

this is simply translated to:

@@

it is evaluated within the Docker container:

@@

At the heart of a PAC lies the planned analysis.
We suggest writing a main analysis function that captures the full process of testing a research question.
This function receives a dataset as input and performs all statistical processing. <!--AB say anything about its output? is the Markdown the output?-->
For the hypothetical research question, we could for example plan to conduct a simple Student's t-test [@studentProbableErrorMean1908] with Welch's correction [@welchGeneralizationStudentProblem1947] an the average of machiavellism items of the Short Dard Triad [SD3, @Jones2013] between the binary-coded gender groups.
As an example of a data-dependent decision, we included a rank transformation if the skewness was greater than 1, converting the t-test to a supposedly more robust Mann–Whitney–Wilcoxon test.[^caution-mww]

[^caution-mww]: The Mann–Whitney–Wilcoxon is in fact **not** more robust than the Welch corrected t-test for reasonably large sample sizes. The interested reader can use the provided code for the simulation to verify that the t-test provides unbiased effect sizes but the Mann–Whitney–Wilcoxon overestimates effect size with increasing sample size and skewness.

This function is one possible translation of the verbal description above:

@@

Importantly, natural language descriptions do not map unambiguously onto computer code implementations. 
The same textual descriptions may be implemented differently by different persons depending on their statistical and programming knowledge but also based on their own expectations and beliefs.
One example would be using the function `wilcox.test` instead of the combinations of `rank` and `t.test`.
Both are valid implementation of the Mann–Whitney–Wilcoxon test but the first assumes equal variance whereas the second applies Welch's correction and hence is robust even with unequal variances across groups [@zimmermanRankTransformationsPower1993].
Mentioning every such minute implementation detail is almost impossible and would result in overly verbose preregistrations.
Still, these details can make a difference and provide unwanted researcher degrees-of-freedom.

Please note that the procedure of running different test procedures based on a fixed skewness threshold was created for didactic purposes and it is not necessarily optimal for adressing your research question.
The t-test and Mann–Whitney–Wilcoxon test are arguably the most often used hypothesis test [@fagerlandTtestsNonparametricTests2012; @hortonStatisticalMethodsJournal2005 reports that 26% of all studies employ a t-test and 27% employ a rank based alternative in the New England Journal of Medicine in 2005].
The analytical strategy presented here is in fact suboptimal with regard to several criteria (untested assumption of measurement invariance [@putnickMeasurementInvarianceConventions2016], underestimation of effect size in presence of measurement error [@frostCorrectingRegressionDilution2000], overestimation of effect size for highly skewed distributions [@stonehouseRobustnessTestsCombined1998]). <!--AB can we move this to a different place? Maybe this would actually better fit to the discussion/limitation section!-->

To use this function one needs a compatible dataset. <!--AB what is "this" function? What does "compatible" mean?-->
At the time of preregistration there is of course no actual data available.
We therefore simulate a dataset first that resembles the expected structure of the empirical data that we will use for the final analysis.

```{r simulate_data}
```
<!--AB it is odd that the return variable is called effect because it is a dataset-->

Here, we decided to randomly draw non-normal data to test how well our procedure performed under non-optimal situations.
Specifically, we draw from a $\chi^2$-distribution, where in one group we add difference that conforms to a given standardized effect size (a Cohen's d of .2).
This simulated dataset fits perfectly into our planned analysis with a sample size of 400 and 10 items: <!--AB in what way? please explain-->

@@

Using the `report` package, we can obtain a preview of how our reporting will look like based on simulated data:

```{r, results='asis', eval=FALSE}
report::report(results$test)
```

This call to the `report()` function will yield the following automatically generated output:

```{r, results='asis', echo=FALSE}
report::report(results$test)
```
<!--AB it is difficult to discern the automatically generated text and out manuscript text even though there is the horizontal line. Not sure how to improve this.-->
***

@@

<!--AB please add a figure number and a proper citation to the figure. Also, tell people where they can find the code to run the power analysis themselves-->

@@

<!--AB: add some text explaining what to once the final data is available-->

# Discussion

<!--AB: TODO add a quick summary -->

# Limitations

<!--AB: add limitations; move the t-test discussion on non-optimality here TODO-->

# Outlook

<!--AB: two lines of what the future of the package, of such standards, and prereigstration is. Maybe say something that preregistration is not always possible or wanted but that reproducibility is almost always a good idea?-->
